Gradual Soundness: Lessons from Static Python

### 2022-04-19 pj-7.1/paper.tex
pj: comment, list of soundness issues
pj: checkpoint for artifact
pj: typed attr restriction

%% TODO after submission, before camera-ready
%% -

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
Its static semantics for types is inspired by optional type checkers; in
particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type serves as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no short-term plans to support all \PEP{} types.
The main priority is to implement a small language of types efficiently.

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
One extra step arises when an untyped method overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

Relative to Python, \SP{} makes a few behavioral changes to enable
sound types and strong optimizations:
class attributes must be set in an \code{\_\_init\_\_} method to
ensure predictable layouts (all \PEP{} checkers agree on this);
module-level variables cannot be mutated; and
multiple inheritance of either typed classes or untyped children
of typed classes is forbidden.

@@

@@

@@

%\acks{
%  \SP{} and the Cinder JIT were developed by
%  Max Bernstein,
%  John Biesnecker,
%  Jacob Bower,
%  Sinan Cepel,
%  Tiansi Hu,
%  Orvid King,
%  Vladimir Matveev,
%  William Meehan,
%  Carl Meyer,
%  Matt Page,
%  Aniket Panse,
%  Brett Simmers,
%  Andrei Talaba,
%  Dino Viehland,
%  and Shiyu Wang.
%  The Brown authors contributed to the language by modeling it independently; they
%  are not part of the \SP{} team.
%
%  Special thanks to Twitter, thanks to which the two teams
%  (specifically, Meyer and Krishnamurthi) met.
%  This work was partly supported by the US National Science Foundation.
%  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
%  TODO partial funded by Meta
%}

>>>>

%% TODO after submission, before camera-ready
%% - reword 'fields' to attribute/descriptor/property?

%% TODO 2022-04-08 from Pyre team
%% Hi Carl! The small notational question I mentioned about in the paper was in
%% Figure 5 -- there's this "M-APP" rule that describes how classmethod invocation
%% is handled. I was trying to figure out how inherited classmethod would work
%% with this rule. Say I have a class B that defines a method f0(B, S0) -> S1, and
%% then I have another class D that inherits B which does not define additional
%% method. Now given Env0 \vdash expr0: S0, how can Env0 \vdash D.f0(expr0): S1 be
%% proven? If I understand correctly, the rule only looks up the type of f0 in
%% Env0(D), i.e. the body of the derived class. But the real definition of f0 is
%% syntactically located in the base class B. Does the notation system used here
%% assume that every function definition in base class gets automatically copied
%% into the derived class? Or should there be an additional preconditions added to
%% the "M-APP" rule saying that f0 is allowed to exist in any of the superclass of
%% D?

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
Its static semantics for types is inspired by optional type checkers; in
particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type serves as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no short-term plans to support all \PEP{} types.

@@

  \item
    Typed fields cannot be overridden by descriptors or properties (\code{@property}).

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
\SP{} can also optimize reads from typed fields.
One extra step arises when an untyped method overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

Relative to Python, \SP{} makes a few behavioral changes to enable
sound types and strong optimizations:
class attributes must be set in an \code{\_\_init\_\_} method to
ensure predictable layouts (all \PEP{} checkers agree on this);
typed attributes are resolved statically, and thus cannot be overridden with
descriptors or properties;
module-level variables cannot be mutated;
typed fields 
and
multiple inheritance of either typed classes or untyped children
of typed classes is forbidden.

@@

%% Carl 2022-04-07
%% Ok, I completed my review of open issues. Turns out #55 was fixed back in
%% December already, we just never closed it :) I found and closed a couple others
%% in the same situation. We are left with a total of 9 issues opened by you all,
%% 8 by KC and 1 by Ben:
%%
%%  https://github.com/facebookincubator/cinder/issues/created_by/LuKC1024
%%
%%  https://github.com/facebookincubator/cinder/issues/created_by/bennn
%%
%% #61 is trivial, I have a fix up and it should land today.
%%
%% Of the rest, I don’t think any are soundness issues. Most are cases where we
%% could just be stricter or issue better errors statically (#40, #49, #50, #51,
%% #52, #60 – #60 might already be fixed, #50 is about to land). #63 is a small
%% improvement to type refinement. #71 is a pretty nasty issue that prevents use
%% of CheckedDict with dynamic key or value type in certain positions that might
%% take us a while to fix. But I’m comfortable saying that none of these are
%% soundness issues. Let me know if your analysis differs.

@@

%% critical soundness issues:
%% https://github.com/facebookincubator/cinder/issues/55
%% https://github.com/facebookincubator/cinder/issues/53
%% https://github.com/facebookincubator/cinder/issues/39
%% https://github.com/facebookincubator/cinder/issues/36

@@

\acks{
  \SP{} and the Cinder JIT were developed by
  Max Bernstein,
  John Biesnecker,
  Jacob Bower,
  Sinan Cepel,
  Tiansi Hu,
  Orvid King,
  Vladimir Matveev,
  William Meehan,
  Carl Meyer,
  Matt Page,
  Aniket Panse,
  Brett Simmers,
  Andrei Talaba,
  Dino Viehland,
  and Shiyu Wang.
  The Brown authors contributed to the language by modeling it independently; they
  are not part of the \SP{} team.

  Special thanks to Twitter, thanks to which the two teams
  (specifically, Meyer and Krishnamurthi) met.
  This work was partly supported by the US National Science Foundation.
  Greenman acknowledges support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
  TODO partial funded by Meta
}
